A fast algorithm for finding efficient solutions in multi-objective problems with black-box multiextremal criteria

### 2022-01-13 2022/OM&S/paper.tex
almost done

\title{A fast algorithm for finding efficient decisions in multi-objective problems with black-box multiextremal functions}

\author{
\name{Konstantin Barkalov, Vladimir Grishagin and Evgeny Kozinov\thanks{CONTACT Evgeny Kozinov. Email: evgeny.kozinov@itmm.unn.ru}}
\affil{Software Department, Lobachevsky State University, Nizhni Novgorod, Russia}
}

@@

The method ALR described above is a modification of the base information-statistical algorithm \cite{Strongin2000,Sergeyev2013} for minimization functions satisfying the H{\"o}lder condition. The modification consists in introducing new characteristics (\ref{eq:17}) after initial stage of the search to improve the method's speedup on account of increasing values of characteristics for intervals containing current record (\ref{eq:18}).

@@

@@

During all experiments the method ALR was used with parameters $r=5.6$, $\sigma = 18$, $T=3$. 
\textcolor[rgb]{1,0,0}{The use of $T=3$ corresponds to alternation of one ``local'' iteration and two ``global'' iterations.}
Accuracy $\varepsilon$ from the stopping condition (\ref{eq:23}) was $0.05$ in bi-criteria problems and $0.01$ in 4-criteria ones.

@@

\textcolor[rgb]{1,0,0}{The Pareto front in one of test problems is shown in Fig.~\ref{fig2a} and Fig.~\ref{fig2b}. Here Fig.~\ref{fig2a} corresponds the Pareto front computed on the uniform grid with the step 0.05 in the search domain whereas Fig.~\ref{fig2b} shows the Pareto front computed by RALR-DA. In the last case it is required only 350 trials to obtain the solution with the same accuracy.}

@@

>>>>

\title{A fast algorithm for finding efficient solutions in multi-objective problems with black-box multiextremal criteria}

\author{
\name{Konstantin Barkalov, Vladimir Grishagin and Evgeny Kozinov\thanks{CONTACT Konstantin Barkalov. Email: konstantin.barkalov@itmm.unn.ru}}
\affil{Software Department, Lobachevsky State University, Nizhni Novgorod, Russia}
}

@@

The method ALR described above is a modification of the basic information-statistical algorithm \cite{Strongin2000,Sergeyev2013} for minimization functions satisfying the H{\"o}lder condition. The modification consists in introducing new characteristics (\ref{eq:17}) after initial stage of the search to improve the method's speedup on account of increasing values of characteristics for intervals containing current record (\ref{eq:18}). 
\textcolor[rgb]{1,0,0}{Actually, the second term in (\ref{eq:17}) will have the largest value equal to $(1.5)^{\sigma}$ only for those intervals for which the equality $v_q = \varphi^*(n)$ or $v_{q-1} = \varphi^*(n)$ holds, i.e. for the intervals with the boundary points correspond to the current record $\varphi^*(n)$.}

@@

\textcolor[rgb]{1,0,0}{The complete convergence theory of the information-statistical algorithm is given in \cite{Strongin2000,Sergeyev2013}.}

@@

During all experiments the method ALR was used with parameters $r=5.6$, $\sigma = 18$, $T=3$. 
\textcolor[rgb]{1,0,0}{The use of $T=3$ corresponds to alternation of one ``local'' iteration and two ``global'' iterations, that can be interpreted as a combination of local refinement and global updating of the current optimizer.}
Accuracy $\varepsilon$ from the stopping condition (\ref{eq:23}) was $0.05$ in bi-criteria problems and $0.01$ in 4-criteria ones.

@@

\textcolor[rgb]{1,0,0}{The Pareto front in one of test problems is shown in Fig.~\ref{fig2a} and Fig.~\ref{fig2b}. Here Fig.~\ref{fig2a} corresponds to the Pareto front computed on the uniform grid with the step 0.05 in the search domain whereas Fig.~\ref{fig2b} shows the Pareto front computed by RALR-DA. In the last case it is required only 608 trials to obtain the solution with the same accuracy.}

@@

\section{Conclusion}

In the paper the multidimensional problems of multicriteria (multiobjective) optimization with black-box time-consuming criteria satisfying the Lipschitz condition are considered. These assumptions lead to significant complexity of the problems under investigation and require the development of special efficient numerical methods for their solving. For this goal in the paper a new algorithm is proposed and its computational scheme is described. The algorithm is based on reducing the MCO problem to a family of single-criterion (scalar) optimization problems generating Pareto-optimal solutions via building parameterized convolutions. In turn, each multidimensional scalar problem is reduced to a univariate one by means of Peano curves that map the multidimensional feasible domain into the real axis. Under Lipschitz condition for criteria the objective functions of the reduced one-dimensional problems satisfy the H{\"o}lder condition and for solving these problems it is necessary to apply specialized global optimization methods. As such method we propose a new algorithm that combines in its decision rule iteration with ``local'' and ``global'' goals. Global iterations provide the convergence to the global optimum and the local ones accelerate the search. Moreover, while solving the univariate problems generated by the family of convolutions the proposed algorithm is able to take into account the criteria values computed already in the course of solving the univariate problems completed earlier. This feature reduces significantly number of trials (calculations of criteria values): the later a problem is solved, the less new criteria computations are produced.

The quality of the proposed algorithm is studied in experiment on 3 representative sets of multiextremal MCO problems of different dimensions and criteria number in comparison with 5 particle swarm and genetic methods. As a criterion of methods' efficiency the hypervolume index estimating the quality of Pareto set approximation is taken. Among all competitors the proposed algorithm demonstrates the best results.

As the directions of further development of the current research it would be interesting to investigate the possibilities of applying other schemes of dimensionality reduction, for example,
\textcolor[rgb]{1,0,0}{the nested optimization scheme \cite{Grishagin2016,Grishagin2018}, the diagonal \cite{Sergeyev2015,Sergeyev2017} or the simplicial \cite{Zilinskas2011,PaulaviciusZilinskas2014} partitions methods}
in the framework of the general approach used in the paper. Moreover, the prospective way is to develop parallel versions of the proposed general scheme oriented at powerful supercomputer systems.
